---
title: "Regression Assignment: Car Miles per Gallon Analysis"
author: "JB"
date: "October 26, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Executive Summary
This assignment will evaluate the impact of different design and property variables 
of cars in the **mtcars** dataset on miles per gallon (mpg).  The evaluation 
will assess the impact of transmission type (am -automatic vs. manual) and other 
variables on mpg with an exploratory analysis, regression and hypothesis testing.

### Exploratory Analysis, Hypothesis Test and Regression
The plot of Figure 1 in the Appendix shows an inital assessment of mpg vs. transmission type.
Manual mpg seems somewhat larger qualitatitively than automatic mpg. A two-tailed, 
two-sample student's t-test will be used to investigate this question further, comparing mpg
for automatic vs. manual transmission.  
```{r hypothesis_test, results = 'hide'}
car_data <- mtcars
t_test <-t.test(data=car_data, mpg~am)
```

```{r pvalue}
t_test$p.value
```
From this result, p = 0.001 implies that there could be a difference in mean 
mpg between the transmission types, assuming a significance level of $\alpha$=0.05.

A linear regression of the mpg vs. transmission was also done to investigate the 
strength of the relationship between these two variables.
``` {r linear}
model_transmission <- lm(mpg~am, data = car_data)
summary_lm <- summary(model_transmission)
print(summary_lm$r.squared)
```
The results show that the correlation between mpg and transmission types seems weak 
(R^2^ = 0.36) and that only about 36%  of the variation in mpg can be explained by 
this model (see Figure 2 in Appendix).

To improve the regression, a multivariate regression model will be done to include additional 
regressor variables beyond transmission.  The following code shows the correlation between 
different regressor variables.  The code determines the correlation between mpg (or weight) 
and every other regressor.
```{r corr}
cor(mtcars)[1,]
cor(mtcars)[6,]
```
From these results, mpg appears highly negatively correlated with weight, disp (engine 
volume),cyl (number of cylinders) and hp (horsepower). However, weight is highly correlated 
with cyl, disp and hp so adding these variables in addition to weight might not add all 
that much value to the model. qsec (1/4 mile time) is very weakly correlated with weight 
and could help the model.

Here are some examples of expanding the model (without intercepts to avoid
am factor split) to include more variables based on these arguments. To save space
the summary of only the 3rd model will be printed.
```{r multivariate regression models}
mr_model <- update(model_transmission,mpg~am+wt-1, data = car_data)
mr_model2 <- update(model_transmission, mpg~am+wt+qsec-1, data=car_data)
mr_model3 <- update(model_transmission, mpg~am+wt+qsec+disp-1, data=car_data)

summary_mr <- summary(mr_model)
summary_mr2 <- summary(mr_model2)
summary_mr3 <- summary(mr_model3)
print(summary_mr3$coefficients)
print(summary_mr3$r.squared)
```
From these results, the regression model is improved by the addition of the wt
and qsec variables (R^2^ = 0.987).  The p-values for each coefficient show 
significance and they are unlikely to be zero.  Adding the disp variable and coefficient
did not really improve the model, and this coefficient did not show significance 
to be different than zero. This was the case when trying the other remaining 
variables as well beyond wt and qsec.  

The analysis of variance (ANOVA) was also used to compare models. The ANOVA p-values show
that adding the weight and qsec variables help to improve the model but disp is not 
adding value beyond this. Other variables to replace disp also did not provide improvement.
Figure 3 shows the residuals of mr_model2. There is no obvious pattern for variance of the
residuals with predicted values and they appear to follow a normal distribution.
```{r anova}
anova_results <- anova(mr_model,mr_model2, mr_model3)
cat("Model 2(am+wt+qsec): ",anova_results$`Pr(>F)`[2])
cat("Model 3(am+wt+qsec+disp): ",anova_results$`Pr(>F)`[3])
```

### Conclusion
There does appear to be a difference in mean mpg with transmission type, with manual
transmission delivering about 7.3 more mean mpg.  However, mpg is not well-predicted by
transmission type alone; adding weight and 1/4 mile time help to improve
the predicted mpg.  The weight coefficient is negative, implying that more weight
reduces the mpg, while the 1/4 mile time coefficient is positive, showing that the
longer time to reach a 1/4 mile could be related to improved fuel efficiency.

### Appendix
**Figure 1:**  Exploratory box plot of miles per galloon for automatic and manual
transmission types.  
```{r exploratory plot, fig.height = 4}
boxplot(data= car_data,mpg~am, main="Car Mileage vs. Transmission Type",
xlab="Transmission Type", ylab="Miles Per Gallon",
col = c("gold"), names=c("Automatic", "Manual"))
```

**Figure 2:**  Linear regression plot of MPG vs transmission type and the associated residuals plot.
```{r linear plot}
par(mfrow=c(1,2))
plot_lm <- plot(car_data$am, car_data$mpg, xlab = "Transmission Type",
ylab = "Miles per gallon", pch = 21, bg = "lightgreen", col = "black",xaxt = 'n',cex=1.5)
axis(side = 1, at = c(0,1),labels = c("Automatic", "Manual"))
abline(model_transmission, lwd = 2)
 
resid_plot <- plot(car_data$am, resid(model_transmission), xlab = "Transmission Type", 
ylab = "Residuals", pch = 21, 
bg = "lightgreen", col = "black", xaxt = 'n', cex=1.5)
axis(side = 1, at = c(0,1),labels = c("Automatic", "Manual"))
abline(h=0, lwd = 2)
```

**Figure 3:**  Residuals plot for the multivariate regression including am, weight and qsec
(Model 2).
```{r multivariate residuals plot,fig.height = 6}
par(mfrow = c(2,2))
plot(mr_model2)
```


